---
title: "Unemployment Insurance County-Level"
author: "PHIOS CUEpidemiology"
date: "05/02/2022"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)
library(lubridate)
library(here)

```

# Create dataframe template with states, quarters, and years

```{r}
states <- read_excel(here("States.xlsx"))

df_template <- bind_rows(ST = rep(states$ST, each = 32),
                         STATE = rep(states$STATE, each = 32),
                         QUARTER = rep(c(1, 2, 3, 4), 408),
                         YEAR = rep(rep(seq(2014, 2021, 1), each = 4), 51))
```


# Read in maximum benefit information and reformat as long data

```{r}
max_wba <- read_excel(here("MaxWBA_2014_2021.xlsx")) |>
  pivot_longer(cols = !ST,
               names_to = c("QUARTER", "YEAR"),
               names_sep = "_",
               names_prefix = "WBAMax_",
               names_transform = list(QUARTER = as.integer, YEAR = as.integer),
               values_to = "MAX_WBA")
```

```{r county hh income,} 
cty_income <- read_csv(here("median_household_income_county.csv"))

cty_income <- cty_income |>
  mutate(state_fips = str_sub(geo_id, start = 1, end = 2), 
         ST = case_when(state_fips == "01" ~ "AL", 
                        state_fips == "02" ~ "AK", 
                        state_fips == "04" ~ "AZ", 
                        state_fips == "05" ~ "AR", 
                        state_fips == "06" ~ "CA", 
                        state_fips == "08" ~ "CO",
                        state_fips == "09" ~ "CT",
                        state_fips == "10" ~ "DE",
                        state_fips == "11" ~ "DC",
                        state_fips == "12" ~ "FL",
                        state_fips == "13" ~ "GA",
                        state_fips == "15" ~ "HI",
                        state_fips == "16" ~ "ID",
                        state_fips == "17" ~ "IL",
                        state_fips == "18" ~ "IN",
                        state_fips == "19" ~ "IA",
                        state_fips == "20" ~ "KS",
                        state_fips == "21" ~ "KY",
                        state_fips == "22" ~ "LA",
                        state_fips == "23" ~ "ME",
                        state_fips == "24" ~ "MD",
                        state_fips == "25" ~ "MA",
                        state_fips == "26" ~ "MI",
                        state_fips == "27" ~ "MN",
                        state_fips == "28" ~ "MS",
                        state_fips == "29" ~ "MO",
                        state_fips == "30" ~ "MT",
                        state_fips == "31" ~ "NE",
                        state_fips == "32" ~ "NV",
                        state_fips == "33" ~ "NH",
                        state_fips == "34" ~ "NJ",
                        state_fips == "35" ~ "NM",
                        state_fips == "36" ~ "NY",
                        state_fips == "37" ~ "NC",
                        state_fips == "38" ~ "ND",
                        state_fips == "39" ~ "OH",
                        state_fips == "40" ~ "OK",
                        state_fips == "41" ~ "OR",
                        state_fips == "42" ~ "PA",
                        state_fips == "44" ~ "RI",
                        state_fips == "45" ~ "SC",
                        state_fips == "46" ~ "SD",
                        state_fips == "47" ~ "TN",
                        state_fips == "48" ~ "TX",
                        state_fips == "49" ~ "UT",
                        state_fips == "50" ~ "VT",
                        state_fips == "51" ~ "VA",
                        state_fips == "53" ~ "WA",
                        state_fips == "54" ~ "WV",
                        state_fips == "55" ~ "WI",
                        state_fips == "56" ~ "WY",
                        state_fips == "60" ~ "AS",
                        state_fips == "66" ~ "GU",
                        state_fips == "69" ~ "MP",
                        state_fips == "72" ~ "PR",
                        state_fips == "78" ~ "VI")) |>
  rename_all(.funs = toupper)

### missing data from https://fred.stlouisfed.org/series/MHINM35039A052NCEN https://datausa.io/profile/geo/loving-county-tx/#economy
cty_income <- cty_income |>
                mutate(MEDIAN_HH_INCOME = ifelse(YEAR == 2018 & GEO_ID == "35039", 41511, 
                                                 ifelse(YEAR == 2015 & GEO_ID == "48301", 81875, 
                                                        ifelse(YEAR == 2020 & GEO_ID == "48243", 55718, MEDIAN_HH_INCOME))))

```

# Combine income and max benefit data

- We're assuming there are 13 weeks in a quarter
- Median income is reported annually

```{r}
ui1 <- left_join(df_template, max_wba, by = c("ST", "QUARTER", "YEAR")) |>
  left_join(cty_income, by = c("ST", "YEAR")) |>
  relocate(STATE_FIPS, GEO_ID, NAME, .before = QUARTER) |>
  mutate(INCOME_QUARTERLY_CTY = MEDIAN_HH_INCOME / 4,
         INCOME_WEEKLY_CTY = MEDIAN_HH_INCOME / 52)
```



# Create variables for calculating weekly benefits allowance (WBA)

Formulas do not change over time for almost all states, so we are manually building in the formulas to the code below. Per Dr. Kate Sacks' work, we are assuming constant income for single individuals with no dependents, full-time work week of 40 hours, and 13 weeks in each quarter. Rounding rules specified in state provisions are incorporated into the formulas.

NOTES: 
- The following states have changes in their WBA formula over time - NY, RI, and SC. 
- Alaska, California, Pennsylvania and West Virginia provide tables of WBA based on ranges of earnings. These values have been input manually.
    - Alaska: https://labor.alaska.gov/unemployment/documents/Calculating_Benefits.pdf, based on annual earnings
    - California: https://www.edd.ca.gov/pdf_pub_ctr/de1101bt5.pdf, based on quarterly earnings
    - Pennsylvania: https://www.uc.pa.gov/unemployment-benefits/Am-I-Eligible/financial-charts/Pages/default.aspx, based on quarterly earnings
    - West Virginia: http://code.wvlegislature.gov/21A-6-10/, based on quarterly earnings

```{r}
ui2 <- ui1 |>
  mutate(WBA_INCOME_CALC = case_when(ST == "AL" ~ ceiling(INCOME_QUARTERLY_CTY / 26),
                                     ST == "AK" ~ 370,
                                     ST == "AZ" ~ round(INCOME_QUARTERLY_CTY / 25),
                                     ST == "AR" ~ floor(INCOME_QUARTERLY_CTY / 26),
                                     ST == "CA" & YEAR == 2014 ~ 377,
                                     ST == "CA" & YEAR == 2015 ~ 383,
                                     ST == "CA" & YEAR == 2016 ~ 394,
                                     ST == "CA" & YEAR == 2017 ~ 395,
                                     ST == "CA" & YEAR == 2018 ~ 408,
                                     ST == "CA" & YEAR == 2019 ~ 425,
                                     ST == "CA" & YEAR == 2020 ~ 450,
                                     ST == "CO" ~ floor(0.6*((INCOME_QUARTERLY_CTY*2) / 26)),
                                     ST == "CT" ~ floor(INCOME_QUARTERLY_CTY / 26),
                                     ST == "DE" ~ floor((INCOME_QUARTERLY_CTY * 2) / 46),
                                     ST == "DC" ~ floor(INCOME_QUARTERLY_CTY / 26),
                                     ST == "FL" ~ floor(INCOME_QUARTERLY_CTY / 26),
                                     ST == "GA" ~ floor((INCOME_QUARTERLY_CTY * 2) / 42),
                                     ST == "HI" ~ ceiling(INCOME_QUARTERLY_CTY / 21),
                                     ST == "ID" ~ floor(INCOME_QUARTERLY_CTY / 26),
                                     ST == "IL" ~ ceiling(0.47 * INCOME_WEEKLY_CTY),
                                     ST == "IN" ~ floor(0.47 * INCOME_WEEKLY_CTY),
                                     ST == "IA" ~ floor(INCOME_QUARTERLY_CTY / 23),
                                     ST == "KS" ~ floor(0.0425 * INCOME_QUARTERLY_CTY),
                                     ST == "KY" ~ round(0.011923 * (INCOME_QUARTERLY_CTY * 4)),
                                     ST == "LA" ~ floor((INCOME_QUARTERLY_CTY / 25) * 1.05 * 1.15),
                                     ST == "ME" ~ floor(INCOME_QUARTERLY_CTY / 22),
                                     ST == "MD" ~ ceiling(INCOME_QUARTERLY_CTY / 24),
                                     ST == "MA" ~ floor(0.5 * INCOME_WEEKLY_CTY),
                                     ST == "MI" ~ floor(0.041 * INCOME_QUARTERLY_CTY),
                                     ST == "MN" ~ floor(0.5 * (INCOME_QUARTERLY_CTY / 13)),
                                     ST == "MS" ~ floor(INCOME_QUARTERLY_CTY / 26),
                                     ST == "MO" ~ floor(0.04 * INCOME_QUARTERLY_CTY),
                                     ST == "MT" ~ floor(0.01 * (INCOME_QUARTERLY_CTY * 4)),
                                     ST == "NE" ~ 2 * floor((0.5 * INCOME_WEEKLY_CTY) / 2),
                                     ST == "NV" ~ floor(INCOME_QUARTERLY_CTY / 25),
                                     ST == "NH" ~ round(0.011 * (INCOME_QUARTERLY_CTY * 4)),
                                     ST == "NJ" ~ floor(0.6 * INCOME_WEEKLY_CTY),
                                     ST == "NM" ~ floor(0.535 * INCOME_WEEKLY_CTY),
                                     ST == "NY" & YEAR < 2020 ~ floor(INCOME_QUARTERLY_CTY / 26),
                                     ST == "NY" & YEAR >= 2020 ~ floor(INCOME_QUARTERLY_CTY / 25),
                                     ST == "NC" ~ floor((2 * INCOME_QUARTERLY_CTY) / 52),
                                     ST == "ND" ~ floor((2.5 * INCOME_QUARTERLY_CTY) / 65),
                                     ST == "OH" ~ floor(0.5 * INCOME_WEEKLY_CTY),
                                     ST == "OK" ~ floor(INCOME_QUARTERLY_CTY / 23),
                                     ST == "OR" ~ floor(0.0125 * (INCOME_QUARTERLY_CTY * 4)),
                                     ST == "PA" & YEAR == 2014 ~ 350,
                                     ST == "PA" & YEAR == 2015 ~ 356,
                                     ST == "PA" & YEAR == 2016 ~ 361,
                                     ST == "PA" & YEAR == 2017 ~ 369,
                                     ST == "PA" & YEAR == 2018 ~ 378,
                                     ST == "PA" & YEAR == 2019 ~ 389,
                                     ST == "PA" & YEAR == 2020 ~ 411,
                                     ST == "RI" & YEAR == 2014 & QUARTER == 1 ~ floor(0.0415 * INCOME_QUARTERLY_CTY),
                                     ST == "RI" ~ floor(0.0385 * INCOME_QUARTERLY_CTY),
                                     ST == "SC" ~ floor(0.5 * INCOME_WEEKLY_CTY),
                                     ST == "SD" ~ floor(INCOME_QUARTERLY_CTY / 26),
                                     ST == "TN" ~ floor(INCOME_QUARTERLY_CTY / 26),
                                     ST == "TX" ~ round(INCOME_QUARTERLY_CTY / 25),
                                     ST == "UT" ~ floor((INCOME_QUARTERLY_CTY / 26) - 5),
                                     ST == "VT" ~ floor((2 * INCOME_QUARTERLY_CTY) / 45),
                                     ST == "VA" ~ floor((2 * INCOME_QUARTERLY_CTY) / 50),
                                     ST == "WA" ~ floor(0.0385 * INCOME_QUARTERLY_CTY),
                                     ST == "WV" & YEAR==2014 ~ 311,
                                     ST == "WV" & YEAR==2015 ~ 319,
                                     ST == "WV" & YEAR==2016 ~ 325,
                                     ST == "WV" & YEAR==2017 ~ 333,
                                     ST == "WV" & YEAR==2018 ~ 344,
                                     ST == "WV" & YEAR==2019 ~ 358,
                                     ST == "WV" & YEAR==2020 ~ 376,
                                     ST == "WI" ~ floor(0.04 * INCOME_QUARTERLY_CTY),
                                     ST == "WY" ~ floor(0.04 * INCOME_QUARTERLY_CTY)),
          WBA_INCOME_MAX = if_else(WBA_INCOME_CALC > MAX_WBA, MAX_WBA, WBA_INCOME_CALC))
```

# Merge in COVID supplementary unemployment insurance data

Add in data on when the 600 and 300 dollar COVID supplementary insurance was active in each state. Calculate the final WBA by adding these values to the calculated WBA.

```{r}

covid_ui <- read_excel(here("WBACalcData/COVID_UI.xlsx"))

ui_final <- left_join(ui2, covid.ui, by = c("ST", "STATE", "QUARTER", "YEAR")) |>
  mutate(WBA_INCOME_FINAL = WBA_INCOME_MAX + COVID600 + COVID300,
         WBA_INCOME_RR = round(WBA_INCOME_FINAL / INCOME_WEEKLY_CTY * 100),
         WBA_INCOME_DIV = round(INCOME_QUARTERLY_CTY / WBA_INCOME_FINAL),
         DATE = yq(paste0(YEAR,": ", QUARTER))) |>
  filter(YEAR < 2021)

```
